---
title: "Assignment 5: Data Visualization"
author: "Jackie Fahrenholz"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each 
instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a 
single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox 
in Sakai. Add your last name into the file name 
(e.g., "Fay_A05_DataVisualization.Rmd") prior to submission.

The completed exercise is due on Monday, February 14 at 7:00 pm.



## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse 
and cowplot packages. Upload the NTL-LTER processed data files for nutrients 
and chemistry/physics for Peter and Paul Lakes (use the tidy 
[`NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv`] version) and 
the processed data file for the Niwot Ridge litter dataset (use the [`NEON_NIWO_Litter_mass_trap_Processed.csv`] version). 

2. Make sure R is reading dates as date format; if not change the format to 
date.

```{r}
#1 
#verify working directory
print(getwd())
#load packages
library(tidyverse)
library(cowplot)
#upload datasets #these run off the page because it wouldn't knit otherwise
chem.nutrients <- read.csv("../Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv", stringsAsFactors = TRUE)
litter <- read.csv("../Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv", stringsAsFactors = TRUE)

#2 
#check dates for format
class(chem.nutrients$sampledate)
#not in date so change it
chem.nutrients$sampledate <- as.Date(chem.nutrients$sampledate, 
                                     format = "%Y-%m-%d")
#check it again
class(chem.nutrients$sampledate)
#check other dataset for date format
class(litter$collectDate)
#change this one too
litter$collectDate <- as.Date(litter$collectDate, format = "%Y-%m-%d")
#check it 
class(litter$collectDate)
```
## Define your theme
3. Build a theme and set it as your default theme.

```{r}
#3
#lets build a theme!
mytheme <- theme_classic(base_size = 13) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "right")
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best 
practices for data visualization. Ensure your theme, color palettes, axes, 
and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with 
separate aesthetics for Peter and Paul lakes. Add a line of best fit and color 
it black. Adjust your axes to hide extreme values (hint: change the limits 
using `xlim()` and `ylim()`).

```{r, fig.height = 5, fig.width = 7, warning = FALSE}
#4
#set theme
theme_set(mytheme) 
#plot this
#define x, y and color by lakename
ggplot(chem.nutrients, aes(x = tp_ug, y = po4, color = lakename))+
  geom_point()+
  #line of best fit, colored black
  geom_smooth(method = "lm")+
  #choose your limits; removing anymore on the x-axis wouldn't be
  #representative, despite there being a clump of data between 0-50
  ylim(0,50) +
  xlim(0,130) +
  ylab('Phosphate') +
  xlab('Phosphorus')+
  labs(color = "Lake Name")
```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, 
and (c) TN, with month as the x axis and lake as a color aesthetic. Then, 
create a cowplot that combines the three graphs. Make sure that only one 
legend is present and that graph axes are aligned.

```{r, fig.height = 8, fig.width = 6, warning = FALSE}
#5
#set theme
theme_set(mytheme) 
#make temp boxplot
temp.plot <- ggplot(chem.nutrients, aes(x = month, y = temperature_C))+
  geom_boxplot(aes(color = lakename))+
  xlab('Month')+
  ylab('Temperature(C)')+
  labs(color = "Lake Name")
# TP plot
tp.plot <- ggplot(chem.nutrients, aes(x = month, y = tp_ug))+
  geom_boxplot(aes(color = lakename))+
  xlab('Month')+
  ylab('TP')+
  labs(color = "Lake Name")
# TN plot
tn.plot <- ggplot(chem.nutrients, aes(x = month, y = tn_ug))+
  geom_boxplot(aes(color = lakename))+
  xlab('Month')+
  ylab('TN')+
  labs(color = "Lake Name")

#use cowplot to make them into one figure
plot_grid(temp.plot + theme(legend.position="none"), 
          tp.plot + theme(legend.position="none"), 
          tn.plot  + theme(legend.position="bottom"), 
          nrow = 3, align = 'hv', rel_heights = c(1, 1))
```

Question: What do you observe about the variables of interest over seasons and 
between lakes? 

> Answer: TP and TN were only measured during three months
(6,7,8) while the temeperature was measured during all months.
This may lead us to believe that you can only sample the other
variables during summer months, or that this pollutant is
present during this time based on seasonal activities. 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the 
"Needles" functional group. Plot the dry mass of needle litter by date and 
separate by NLCD class with a color aesthetic. (no need to adjust the name of 
each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into 
three facets rather than separated by color. 

```{r, fig.height = 5, fig.width = 7, warning = FALSE}
#6
#set theme
theme_set(mytheme)
#plot this by subsetting the litter dataset into Needles only
litter.subset <- ggplot(subset(litter, functionalGroup == "Needles"),
                        aes(x = collectDate, y = dryMass,color = nlcdClass)) +
  geom_point() +
  xlab("Date") +
  ylab("Dry Mass") +
  labs(color = "NLCD Class")
print(litter.subset)
#7
#now do the same thing but with a facet instead of by color
litter.facet <- ggplot(subset(litter, functionalGroup == "Needles"), 
                       aes(x = collectDate, y = dryMass)) +
  facet_wrap(vars(nlcdClass), ncol = 3) +
  geom_point() +
  xlab("Date") +
  ylab("Dry Mass")
print(litter.facet)
# lets put them next to each other to see what we think
plot_grid(litter.facet, litter.subset, nrow = 2, align = 'hv', 
          rel_heights = c(1, 1))
```

Question: Which of these plots (6 vs. 7) do you think is more effective, 
and why? 

> Answer: I think that the facetwrwap graph `litter.facet` is more effective in displaying the differences between each of the classes because you can visually 
compare the density of points next to each other as well as the variation 
between years and types of nlcd class across the board. The first graph though 
it's visually appealing doesn't give us as full of a picture but rather 
displaying how similar or different they are based on ncld class in the given 
year. These plots are more crowded though so only general conclusions could be 
made based on this scale. 
